### 2015-05-04 writeup/paper.tex
Heh

%-----------------------------------------------------------------------------
%
%               Template for sigplanconf LaTeX Class
%
% Name:         sigplanconf-template.tex
%
% Purpose:      A template for sigplanconf.cls, which is a LaTeX 2e class
%               file for SIGPLAN conference proceedings.
%
% Guide:        Refer to "Author's Guide to the ACM SIGPLAN Class,"
%               sigplanconf-guide.pdf
%
% Author:       Paul C. Anagnostopoulos
%               Windfall Software
%               978 371-2316
%               paul@windfall.com
%
% Created:      15 February 2005
%
%-----------------------------------------------------------------------------

\documentclass[preprint]{sigplanconf}

% The following \documentclass options may be useful:
%
% 10pt          To set in 10-point type instead of 9-point.
% 11pt          To set in 11-point type instead of 9-point.
% authoryear    To obtain author/year citation style instead of numeric.

\usepackage{amsmath, listings, bcprules, amsthm, amssymb, proof}
\usepackage{textcomp}

\begin{document}

\newtheorem{theorem}{Theorem}[section]
\newtheorem{lemma}[theorem]{Lemma}
\newtheorem{definition}[theorem]{Definition}
\newtheorem{proposition}[theorem]{Proposition}

\conferenceinfo{B629 '15}{Bloomington, IN.}
\copyrightyear{2015} 
\copyrightdata{[to be supplied]}

\lstnewenvironment{code}[1][]%
  {\noindent\medskip 
     \lstset{basicstyle=\sffamily\footnotesize,frame=single,#1}}
  {}

%\titlebanner{banner above paper title}        % These are ignored unless
%\preprintfooter{short description of paper}   % 'preprint' option specified.

\title{Laurens de Graaf}
\subtitle{A Spineless, ``Tagless'' Graph-Reduction Machine in RPython}

%% \title{Semantics for Explicit Contract Monitoring and Support for Multiple Evaluation Strategies}
%\subtitle{Subtitle Text, if any}

\lstset{language=Python, basicstyle=\sffamily}

%% \authorinfo{Name1}
%%            {Affiliation1}
%%            {Email1}
\authorinfo{Cameron Swords}
           {Indiana University}
           {\{cswords\}@indiana.edu}
\maketitle

\begin{abstract}
\begin{center}
\emph{``A Great and Mischievous Pirate''
{\begin{flushright}$\sim$ Henry Morgan\end{flushright}}}
\end{center}
\end{abstract}

%% \category{CR-number}{subcategory}{third-level}
\terms
Languages, JIT, Tracing
\keywords
piracy, rpython, tracing, jit

\section{\emph{Laurencillo}} %% Introduction

There has been tremendous work into JITs and JIT optimizations in recent years.
Most notably, the RPython tracing JIT has come to be a de-facto standard for the
power of a tracing JIT, driving their flagship python interpreter PyPy. In this
work we present an implementation of the Spinless, Tagless G-Machine (herein the
STG Machine) as an interpreter written on top of RPython.

The original aim of this work was to construct a fast implementation of the
STG machine that might support a reasonable subset of the compilation output
of GHC. To this end, our implementation is faithful to the original semantics
presented by \citet{spj:stgmachine}.

\section{\emph{Tigre}} %% The Spineless, Tagless G-Machine

In this section, we give a brief sketch of the implementation approach we take
to construct the STG Machine in RPython, discussing the general structure of the
implementation. To begin, we sketch the directory structure and describe each file
there:

\begin{itemize}
\item \lstinline{laurens/}
  \begin{itemize}
    \item \lstinline{config.py} -- Configuration file (currently unused)
    \item \lstinline{debug.py}  -- Debug utilities, primarily for printing
                                   machine states during execution
    \item \lstinline{op.py}     -- The operation nodes, which each contain
                                   a \lstinline{step} operation that performs
                                   a machine reduction.
    \item \lstinline{parse.py}  -- The sketch of an input parser (currently 
                                   unstarted)
    \item \lstinline{stg.py}    -- The main loop of the implementation
    \item \lstinline{tests.py}  -- A set of tests written in terms of AST
                                   nodes for the machine; this serves as
                                   an alternative entry point
  \item \lstinline{laurens/ast/}
    \begin{itemize}
      \item \lstinline{ast.py}  -- Contains the AST nodes that make up expressions;
                                   part of \lstinline{op.py} should likely be
                                   refactored into this file
      \item \lstinline{cont.py} -- Continuations that are pushed to the Return Stack 
    \end{itemize}
  \item \lstinline{laurens/data/}
    \begin{itemize}
      \item \lstinline{argstack.py} -- Stack for arguments to functions
      \item \lstinline{closure.py}  -- Data structure to represent closures
      \item \lstinline{config.py}   -- Configuration file (currently unused)
      \item \lstinline{heap.py}     -- Heap structure
      \item \lstinline{heapobj.py}  -- Objects that live on the heap
      \item \lstinline{retstack.py} --
      \item \lstinline{updframe.py} --
      \item \lstinline{updstack.py} -- 
    \end{itemize}
  \end{itemize}
\end{itemize}

\section{\emph{Francesca}} %% RPython

In this section, we discuss a number of intricacies that require careful
consideration during implementation of the STG machine.

\subsection{The Configuration Structure}
We begin by briefly sketching the machine configuration, which contains
six individual pieces:
\[\langle op; as; rs; us; h; \sigma\rangle\]
Unsurprisingly, each component plays a unique role in the machine:
\begin{itemize}
  \item $op$ contains the operation for the reduction to perform;
  \item $as$ contains the argument stack, which maintains arguments
        during applications;
  \item $rs$ contains the return stack, which manages return points
        (primarily ``case'' expressions);
  \item $us$ contain the update stack, used for handling thunk updates;
  \item $h$ contains the heap, where memory addresses map to closures;
  \item $\sigma$ contains the \emph{global environment} of bindings.
\end{itemize}
While \citet{spj:stgmachine} observe that many of these stacks may
be combined during compilation to C as an optimization, we maintain
this configuration explicitly in our implementation:
\begin{code}
class Config(object):
  def __init__(self, code, args, ret,  
                     upd, heap, genv):
    self.code       = code
    self.arg_stack  = args
    self.ret_stack  = ret
    self.upd_stack  = upd
    self.heap       = heap
    self.global_env = genv
\end{code}
We take this approach for two reasons. First, we are working from the
assumption that RPython should be able to properly handle the interpretation
structure of the STG machine, and thus these individual stacks should not
cause critical, performance-inhibiting overhead.
Second, RPython performs immense amounts of type inference and combining
these stacks would require that we carefully package each different node
in a new object to indicate which of the original stacks it lived on,
thus adding another level of indirection that the compiler must 
overcome\footnote{There are other mechanisms to avoid this discussed
in the original work, but they require subtle bit manipulation that
would need to be faked in a higher-level implementation.}.

This automatic type distinction has a further problem, however: we need
identical classes for each of the \lstinline{arg_stack}, \lstinline{ret_stack},
and \lstinline{upd_stack} so that RPython may infer each to a different,
appropriate type. Indeed, we would suggest that this is one of the largest
hurdles in constructing a project such as this in RPython: there is a non-zero
amount of code duplication required to simply exist in the RPython ecosystem.

\subsection{The Funny Case of Laziness}

While standard lazy languages provide forcing at primitive operations, the
STG machine departs from this explicit approach. Instead, a primitive operation
assumes that its arguments are pre-evaluated, and this became troublesome when
testing them. To further elaborate this point, observe a standard definition
of addition in the STG language:
\begin{code}[language=Haskell]
+ = \ {} \\n {n,m} -> case n of
                        n' -> case m of
                                m' -> #+ n' m'
\end{code}
This snippet is interpreted as: 
\begin{center}
\emph{
``Plus is a lambda with no free variables that
is should not be updated. When applied, it will \lstinline{case} on
\lstinline{n} and then \lstinline{m}, using the forcing position of
\lstinline{case} to ensure that both inputs are evaluated before using the
primitive addition operator \lstinline{\#+} on the result.''
}
\end{center}
While not particularly tied to the implementation in RPython, we found this
insight invaluable during implementation and testing and so we have included
it here. This manifests with the following dispatch for addition in our
implementation:
\begin{code}
elif expr_type is ast.ast.PrimOp:
  if cexp.oper == "+":
    lookups = vals(code.env,
                   global_env, 
                   cexp.atoms)
    x1      = lookups[0]
    x2      = lookups[1]
    assert isinstance(x1,ast.ast.ValAST)
    assert isinstance(x2,ast.ast.ValAST)
    res     = x1.value + x2.value
    config.code = op.ReturnInt(res) 
\end{code}
We first ensure that we are performing addition and then look up the operators
with \lstinline{vals} (which uses the local and global environments for variable
lookup). Next, we ensure that each of the resulting lookups are valid Value
nodes, extract their values, and then construct a \lstinline{ReturnIn} operation
code for the configuration.

\section{\emph{Neptune} n\'{e}e \emph{Fortune}} %% Conclusion / Future Work

%% \section{Appendix Title}
%% 
%% This is the text of the appendix, if you need one.

%% \acks
%% 
%% Acknowledgments, if needed.

% We recommend abbrvnat bibliography style.

\bibliographystyle{abbrvnat}

% The bibliography should be embedded for final submission.

\softraggedright
\bibliography{p}

%%\appendix
%%\input{proof-appendix.tex}

\end{document}

### 2015-05-04 writeup/paper.tex
Two sigplan pages are done, but the citations are a mess

@@
   {\noindent\medskip 
-     \lstset{basicstyle=\sffamily\footnotesize,frame=single,#1}}
+     %\lstset{basicstyle=\sffamily\footnotesize,frame=single,#1}}
+     \lstset{basicstyle=\sffamily\footnotesize,#1}}
   {}
@@
 implementation. To begin, we sketch the directory structure and describe each file
-there:
+in Figure~\ref{fig:dir} to outline the overall project.
 
+\begin{figure}
 \begin{itemize}
-\item \lstinline{laurens/}
+\item \lstinline{laurens/} -- Top-level directory
   \begin{itemize}
@@
                                    a machine reduction.
-    \item \lstinline{parse.py}  -- The sketch of an input parser (currently 
-                                   unstarted)
+    \item \lstinline{parse.py}  -- Input parser (currently unstarted)
     \item \lstinline{stg.py}    -- The main loop of the implementation
@@
     \begin{itemize}
-      \item \lstinline{ast.py}  -- Contains the AST nodes that make up expressions;
+      \item \lstinline{ast.py}  -- AST nodes that make up expressions;
                                    part of \lstinline{op.py} should likely be
                                    refactored into this file
-      \item \lstinline{cont.py} -- Continuations that are pushed to the Return Stack 
+      \item \lstinline{cont.py} -- Continuations for the  Return Stack 
     \end{itemize}
@@
       \item \lstinline{closure.py}  -- Data structure to represent closures
-      \item \lstinline{config.py}   -- Configuration file (currently unused)
+      \item \lstinline{config.py}   -- Machine Configuration
       \item \lstinline{heap.py}     -- Heap structure
-      \item \lstinline{heapobj.py}  -- Objects that live on the heap
-      \item \lstinline{retstack.py} --
-      \item \lstinline{updframe.py} --
-      \item \lstinline{updstack.py} -- 
+      \item \lstinline{retstack.py} -- Stack for return continuations
+      \item \lstinline{updframe.py} -- Update frame tuple
+      \item \lstinline{updstack.py} -- Stack for update frames
     \end{itemize}
@@
 \end{itemize}
+\caption{\label{fig:dir}The directory structure of Laurens.}
+\end{figure}
 
-\section{\emph{Francesca}} %% RPython
+\subsection{The Machine Configuration}
 
-In this section, we discuss a number of intricacies that require careful
-consideration during implementation of the STG machine.
-
-\subsection{The Configuration Structure}
-We begin by briefly sketching the machine configuration, which contains
-six individual pieces:
+Before we discuss the implementation details, we describe the
+STG Machine. This abstract machine uses a six-tuple configuration:
 \[\langle op; as; rs; us; h; \sigma\rangle\]
@@
 cause critical, performance-inhibiting overhead.
-Second, RPython performs immense amounts of type inference and combining
-these stacks would require that we carefully package each different node
-in a new object to indicate which of the original stacks it lived on,
-thus adding another level of indirection that the compiler must 
-overcome\footnote{There are other mechanisms to avoid this discussed
-in the original work, but they require subtle bit manipulation that
-would need to be faked in a higher-level implementation.}.
+Second, RPython performs immense amounts of type inference and so we use
+separate stack implementations, as we discuss further in
+Section~\ref{sec:typeinf}.
 
-This automatic type distinction has a further problem, however: we need
-identical classes for each of the \lstinline{arg_stack}, \lstinline{ret_stack},
-and \lstinline{upd_stack} so that RPython may infer each to a different,
-appropriate type. Indeed, we would suggest that this is one of the largest
+\subsection{The Main Loop}
+
+\begin{figure}
+\begin{code}
+def terminateHuh(config):
+  return 
+    (((config.code.op == op.ReturnConOp) or 
+      (config.code.op == op.ReturnIntOp)) 
+     and config.ret_stack.empty()
+     and config.upd_stack.empty())
+
+def loop(config):
+  while not terminateHuh(config):
+    jitdriver.jit_merge_point(code = 
+                                config.code)
+    config = config.code.step(config)
+  return config
+\end{code}
+\caption{\label{fig:loop}The main loop and termination test.}
+\end{figure}
+
+The \lstinline{stg.py} file serves as the main loop, driven by operation steps
+in \lstinline{op.py}. Everything else supports this main interaction.  
+Our overall computational mechanism is presented in Figure~\ref{fig:loop}. The
+main file contains a loop driver that steps the configuration until there is no
+more to do. At each step, we call \lstinline{step} on the current configuration,
+receiving a new one. This computation proceeds until we terminate with either
+an integer or a constructor and both the update and return stacks are empty.
+
+The operations are broken up by the STG machine into four possible operations:
+evaluation, entering, returning a construct, and returning an integer, given
+as \emph{Eval $e$ $\rho$}, \emph{Enter $a$}, \emph{ReturnCon $c$ $xs$}, and
+\emph{ReturnInt $i$} in the reduction semantics. We implement each of these
+as an \emph{operation node} in \lstinline{op.py}, wherein each has a
+\lstinline{step} method that expects a configuration using that operator
+and dispatches appropriately, returning the new machine configuration.
+
+\section{\emph{Francesca}} %% RPython
+
+In this section, we discuss a number of intricacies that require careful
+consideration during implementation of the STG machine.
+
+\subsection{Unique Stack Implementations}\label{sec:typeinf}
+
+As previously state, RPython attempts exhaustive type inference for structures
+during compilation. This approach has forced us into using three different
+stack implementations, all identical in code; we would like to use a single
+stack defined as \lstinline{data/stack.py}, but these authors could not find
+a way to circumvent the type inference to allow such reuse.
+Indeed, we would suggest that this is one of the largest
 hurdles in constructing a project such as this in RPython: there is a non-zero
 amount of code duplication required to simply exist in the RPython ecosystem.
+
+Furthermore, the later sections of \citet{spj:stgmachine} describe an
+optimization for combining stacks that would suggest that we can combine these
+stacks by carefully packaging each different node in a new object to indicate
+which of the original stacks it lived on\footnote{There are other mechanisms to
+avoid this discussed in the original work, but they require subtle bit
+manipulation that would need to be faked in a higher-level implementation.}. 
+Unfortunately, this would add one further level of indirection for the compiler
+and further obfuscate our implementation, and it is unclear if we could gain
+speed for the effort.  Further experimentation should be done, here, however.
 
@@
 
+Finally, we discuss our timing results, and must report a negative one:
+the final implementation is \emph{slower} when we perform tracing on
+a sample program computing the 30th Fibonacci number. This is almost certainly
+the result of our na\"{i}ve tracing mechanism, wherien we consider every node as
+a possible merge point for the JIT, introducing immense tracing overhead to the
+computation. Thus it is unclear if Tracing is suitable for the STG machine
+because we have not performed effort to fine-tune our implementation.
+
+Beyond this result, we still have much to do:
+\begin{itemize}
+\item Write a full parser that will construct AST nodes from reasonable input;
+\item Write a more extensive test suite to examine and use constructors,
+      \lstinline{let}, \lstinline{letrec}, laziness, etc.; this will be much
+      easier once a parser is written;
+\item Support a full range of built-in functions, including I/O and additional
+      numeric operators;
+\item Replace the dictionary-based representation of environments with a
+      more efficient data structure
+\item Fine-tune the JIT heuristics to gain speed.
+\end{itemize}
+
 %% \section{Appendix Title}

### 2015-05-04 laurens/stg.py

@@ -31,29 +31,16 @@ jitdriver = JitDriver(greens=['code']
 def terminateHuh(config):
   return (((config.code.op == op.ReturnConOp) or 
            (config.code.op == op.ReturnIntOp)) 
-          and (not config.ret_stack.peek()[0]))
+          and config.ret_stack.empty()
+          and config.upd_stack.empty())
 
 def loop(config):
 
   while not terminateHuh(config):
     jitdriver.jit_merge_point( code       = config.code)
-                            #, config     = config
-                            #, arg_stack  = config.arg_stack 
-                            #, ret_stack  = config.ret_stack 
-                            #, upd_stack  = config.upd_stack 
-                            #, heap       = config.heap
-                            #, global_env = config.global_env)
     debug("-------------")
     debug(str(config))
     config = config.code.step(config)
-
-    #jitdriver.can_enter_jit(code=config.code)
-    #                       , config     = config
-    #                       , arg_stack  = config.arg_stack 
-    #                       , ret_stack  = config.ret_stack 
-    #                       , upd_stack  = config.upd_stack 
-    #                       , heap       = config.heap
-    #                       , global_env = config.global_env)   
   return config
 
 def run(fp):

### 2015-05-06 writeup/paper.tex
More write-up done

@@
 
-\section{\emph{Laurencillo}} %% Introduction
+\section{\emph{Laurencillo}~-- Introduction} %% Introduction
 
@@
 
-\section{\emph{Tigre}} %% The Spineless, Tagless G-Machine
+\section{\emph{Tigre}~-- The STG Machine} %% The Spineless, Tagless G-Machine
 
@@
 evaluation, entering, returning a construct, and returning an integer, given
-as \emph{Eval $e$ $\rho$}, \emph{Enter $a$}, \emph{ReturnCon $c$ $xs$}, and
-\emph{ReturnInt $i$} in the reduction semantics. We implement each of these
+as \emph{Eval $e$ $\rho$}, \emph{Enter $a$}, \emph{ReturnInt $i$}, and
+\emph{ReturnCon $c$ $xs$} in the reduction semantics. We implement each of these
 as an \emph{operation node} in \lstinline{op.py}, wherein each has a
-\lstinline{step} method that expects a configuration using that operator
-and dispatches appropriately, returning the new machine configuration.
+\lstinline{step} method that expects a configuration using that operator and
+dispatches appropriately, returning the new machine configuration.
 
-\section{\emph{Francesca}} %% RPython
+\subsection{Operations}
+
+There are four operations in the STG machine which indicate the next
+step the machine will take. We implement each of these as a \lstinline{step}
+attached to an operation node in \lstinline{op.py}, allowing us to easily
+modularize the main loop.
+
+\paragraph{\textit{Eval e $\rho$}.}
+The \textit{Eval} step evaluates an operation, yielding either a 
+sub-expression to evaluate (in the case of a \lstinline{case} operation)
+or an \textit{Enter} operation with a heap address location (such as when
+evaluating an application).
+This is, by far, the most grotesque part of the implementation: to implement the
+evaluation step, we extract the expression node and branch on its type,
+constructing a new configuration as necessary based on the result.
+
+\paragraph{\textit{Enter a}.}
+The \textit{Enter} step enters a heap address, which is a \emph{closure} stored
+in the heap. This operation also handles thunk updates and partial applications,
+constructing an updated thunk in the former case and a new thunk with additional
+input in the latter.
+
+\paragraph{\textit{ReturnInt i}.}
+This operation indicates that an expression has evaluated to an integer $i$. The
+return stack will contain a \lstinline{case} expression to match against this
+result. We present this case in full in Figure~\ref{fig:ReturnInt}; the 
+\lstinline{int_case_lookup} operator takes the relevant cases and identifies the
+appropriate one. We also indicate a \emph{default case}, which uses a binder
+with a default variable to bind the result if it does not match an integer
+literal.
+
+\begin{figure}
+\begin{code}
+def step(self,config):
+    debug("Int return")
+
+    if config.ret_stack.empty():
+      raise Exception('...')
+    else: 
+      retk      = config.ret_stack.pop()
+      if type(retk) is ast.cont.CaseCont:
+        ret_env   = retk.env.copy()
+        ret_alts  = retk.alts
+        value     = self.value
+
+        default,default_var,body = 
+          self.int_case_lookup(value, ret_alts)
+
+        if default:
+          if (default_var is not None):
+            var = default_var
+            ret_env[var] = 
+              ast.ast.Value(value, True)
+
+        config.code = op.Eval(body, ret_env)
+
+    return config
+
+  def int_case_lookup(self,val,ret_alts):
+    # Assuming everything is a LitAlt
+    for alt in ret_alts.alternates: 
+      if alt.literal == val:
+        return (False, None, alt.rhs)
+
+    return (True, 
+            ret_alts.default.binder, 
+            ret_alts.default.rhs)
+\end{code}
+\caption{\label{fig:returnint}The \textit{ReturnInt} operation.}
+\end{figure}
+
+\paragraph{\textit{ReturnCon c xs}.}
+This operation indicates that an expression has evaluated to a constructor $c$
+with arguments in the $xs$. The return stack (or, if it is empty, the update
+stack) will contain a \lstinline{case} expression to match against this result.
+
+\section{\emph{Francesca}~-- RPython Usage \& Notes} %% RPython
 
 In this section, we discuss a number of intricacies that require careful
-consideration during implementation of the STG machine.
+consideration during implementation of the STG machine before describing
+how we have leveraged the \lstinline{jitdriver} in our implementation.
 
-\subsection{Unique Stack Implementations}\label{sec:typeinf}
+\subsection{RPython Customizations}
 
+\paragraph{Unique Stack Implementations.}\label{sec:typeinf}
 As previously state, RPython attempts exhaustive type inference for structures
@@
 
-Furthermore, the later sections of \citet{spj:stgmachine} describe an
+Furthermore, the machine description~\cite{spj:stgmachine} describe an
 optimization for combining stacks that would suggest that we can combine these
@@
 
-\subsection{The Funny Case of Laziness}
-
+\paragraph{The Funny Case of Laziness.}
 While standard lazy languages provide forcing at primitive operations, the
@@
 
-\section{\emph{Neptune} n\'{e}e \emph{Fortune}} %% Conclusion / Future Work
+\subsection{The RPython JIT Driver.}
+
+\begin{code}
+from rpython.rlib.jit import JitDriver, purefunction
+
+def get_location(code):
+    return "%s" % ( code )
+
+jitdriver = JitDriver(greens=['code']
+                     , reds='auto'
+                     , get_printable_location=get_location)
+
+def terminateHuh(config):
+  return (((config.code.op == op.ReturnConOp) or 
+           (config.code.op == op.ReturnIntOp)) 
+          and config.ret_stack.empty()
+          and config.upd_stack.empty())
+
+def loop(config):
+
+  while not terminateHuh(config):
+    jitdriver.jit_merge_point( code       = config.code)
+    config = config.code.step(config)
+  return config
+\end{code}
+
+\section{\emph{Neptune} n\'{e}e \emph{Fortune}~-- Conclusion} %% Conclusion / Future Work
 

### 2015-05-06 writeup/paper.tex
Heh

@@
 
-\usepackage{amsmath, listings, bcprules, amsthm, amssymb, proof}
+\usepackage{amsmath, listings, amsthm, amssymb, proof}
 \usepackage{textcomp}
@@
 There has been tremendous work into JITs and JIT optimizations in recent years.
-Most notably, the RPython tracing JIT has come to be a de-facto standard for the
-power of a tracing JIT, driving their flagship python interpreter PyPy. In this
-work we present an implementation of the Spinless, Tagless G-Machine (herein the
-STG Machine) as an interpreter written on top of RPython.
+Most notably, the RPython tracing JIT~\cite{Bolz:2009} has come to be
+a de-facto standard for the power of a tracing JIT, driving their
+flagship python interpreter PyPy. In this work we present an
+implementation of the Spinless, Tagless
+G-Machine~\cite{spj:stgmachine} (herein the STG Machine) as an
+interpreter written on top of RPython.
 
@@
 
+After constructing an STG machine interpreter that functioned in
+\lstinline{PyPy}, we spent notable work fixing a number of
+implementation details to work with the RPython compiler. These
+optimizations include data structure duplication due to type
+constraints, replacing \lstinline{==} with \lstinline{is}
+for efficiency, and using \lstinline{assert} and subclassing
+to aid in type-directed optimizations.
+
 \paragraph{Unique Stack Implementations.}\label{sec:typeinf}
@@
 
-\paragraph{The Funny Case of Laziness.}
+\paragraph{Primitive Operations and Assertions.}
 While standard lazy languages provide forcing at primitive operations, the
@@
 \end{center}
-While not particularly tied to the implementation in RPython, we found this
-insight invaluable during implementation and testing and so we have included
-it here. This manifests with the following dispatch for addition in our
-implementation:
+This approach to implementing standard primitive operators yielded an
+additional problem in RPython: during environment lookup, the
+inference system had no recourse to determine that our resultant lookup
+values were, in fact, integers. To this end, we added an additional
+class to the AST class hierarchy, a \lstinline{ValAST}, and added
+\lstinline{assert} statements during primitive operator evaluation
+facilitate value extraction:
 \begin{code}

### 2015-05-06 test/Fib.hs
Added haskell fib test

@@ -1,0 +1,7 @@
+module Fib where
+
+fib 0 = 0
+fib 1 = 1
+fib n = fib (n - 1) + fib (n - 2)
+
+main = fib 30

### 2015-05-06 writeup/paper.tex
Write up is done

@@
 {\begin{flushright}$\sim$ Henry Morgan\end{flushright}}}
+\emph{``I didn't proof-read this paper before I turned it in. I'm certain it's rough,
+       and needs at least one more rewrite.''
+{\begin{flushright}$\sim$ Cameron Swords\end{flushright}}}
 \end{center}
@@
   while not terminateHuh(config):
-    jitdriver.jit_merge_point(code = 
-                                config.code)
     config = config.code.step(config)
@@
 return stack will contain a \lstinline{case} expression to match against this
-result. We present this case in full in Figure~\ref{fig:ReturnInt}; the 
+result. We present this case in full in Figure~\ref{fig:returnint}; the 
 \lstinline{int_case_lookup} operator takes the relevant cases and identifies the
@@
 
+We implement a na\"{i}ve JIT driver with RPthon in our main loop, modifying it
+to support a merge point and identifying the \lstinline{code} piece of the 
+configuration as the \emph{green} portion. This modified main loop is presented
+in Figure~\ref{fig:jitloop}.
+\begin{figure}
 \begin{code}
-from rpython.rlib.jit import JitDriver, purefunction
+from rpython.rlib.jit 
+  import JitDriver, purefunction
 
@@
 
-jitdriver = JitDriver(greens=['code']
+jitdriver = 
+  JitDriver( greens=['code']
            , reds='auto'
-                     , get_printable_location=get_location)
+           , get_printable_location =
+               get_location)
 
@@
   while not terminateHuh(config):
-    jitdriver.jit_merge_point( code       = config.code)
+    jitdriver.jit_merge_point( code = 
+                                 config.code)
     config = config.code.step(config)
@@
 \end{code}
+\caption{\label{fig:jitloop}The JIT-sensitive main loop.}
+\end{figure}
+
+\subsection{Timing Results}
+
+Our timing results are surprisingly poor: 
+the final implementation is \emph{slower} when we perform tracing on
+a sample program computing the 30th Fibonacci number, as opposed to running
+it through PyPy (at 18 seconds versus 14 seconds). Perhaps less shocking,
+GHC can compile and run an equivalent program in approximately 0.022 seconds,
+suggesting we are several \emph{orders of magnitude} slower than the current
+GHC implementation.
+
+This is almost certainly the result of our na\"{i}ve tracing mechanism, wherein
+we consider every node as a possible merge point for the JIT, introducing
+immense tracing overhead to the computation. The first optimization we should
+explore is only considering \textit{Enter} operations for merge points. The
+second, less clear optimization, is that of termination conditions: using a
+termination condition as we have in our implementation can be hard on the
+tracing mechanism, but it is unclear without further investigation how our STG
+implementation might be modified to ``politely'' terminate, given the continuous
+interaction of the return and update stacks.
 
@@
 
-Finally, we discuss our timing results, and must report a negative one:
-the final implementation is \emph{slower} when we perform tracing on
-a sample program computing the 30th Fibonacci number. This is almost certainly
-the result of our na\"{i}ve tracing mechanism, wherien we consider every node as
-a possible merge point for the JIT, introducing immense tracing overhead to the
-computation. Thus it is unclear if Tracing is suitable for the STG machine
-because we have not performed effort to fine-tune our implementation.
+In closing, we found this project enlightening and insightful: the idea of a 
+tracing JIT is appealing, but there are a number of difficulties and concerns
+that arise in the area almost immediately that we have tried to discuss in this
+paper. Tracing seems, in some ways, a brittle box---at least in RPython, there
+are a number of hoops to leap through to get code compiled. To this end,
+it would would be interesting to see a similar system built in a strongly-typed
+language in order to avoid the ``surprise'' compilation errors that one
+experiences when building facilities in RPython. Compared to the standard
+approach of building a compiler, however, this alternative has a lot of promise
+in terms of rapid prototyping, language evolution, and general lack of
+compiler-specific expert knowledge. While the STG machine may have been a poor
+choice for implementation (due to the authors' lack of familiarity with its
+intricacies), using a system like RPython in the future for quick prototyping
+is certainly reasonable and worthwhile.
 
-Beyond this result, we still have much to do:
+Beyond this, we still have much to do:
 \begin{itemize}
@@
       more efficient data structure
+\item Consider breaking up the \textit{Eval} implementation into individual
+      AST nodes.
 \item Fine-tune the JIT heuristics to gain speed.
@@
 
-%% \acks
-%% 
-%% Acknowledgments, if needed.
+\acks
+
+We'd like to thank Spenser Bauman for his immense interest and help
+in this work, and Sam Tobin-Hochstadt for facilitating the creation
+of this project as part of his Metaprogramming course.
 

### 2015-05-06 stg.py
Delete stg.py

@@ -1,145 +1,0 @@
-"""
-PyPy tutorial by Andrew Brown
-example5.py - BF interpreter in RPython, translatable by PyPy, with JIT, and
-              with a get_printable_location function and pure function wrapper
-              for the dictionary lookup.
-
-"""
-
-import os
-import sys
-
-# So that you can still run this module under standard CPython, I add this
-# import guard that creates a dummy class instead.
-# try:
-from rpython.rlib.jit import JitDriver, purefunction
-# except ImportError:
-#     class JitDriver(object):
-#         def __init__(self,**kw): pass
-#         def jit_merge_point(self,**kw): pass
-#         def can_enter_jit(self,**kw): pass
-#     def purefunction(f): return f
-
-def get_location(pc, program, bracket_map):
-    return "%s_%s_%s" % (
-            program[:pc], program[pc], program[pc+1:]
-            )
-
-jitdriver = JitDriver(greens=['pc', 'program', 'bracket_map'], reds=['tape'],
-        get_printable_location=get_location)
-
-@purefunction
-def get_matching_bracket(bracket_map, pc):
-    return bracket_map[pc]
-
-def mainloop(program, bracket_map):
-    pc = 0
-    tape = Tape()
-    
-    while pc < len(program):
-        jitdriver.jit_merge_point(pc=pc, tape=tape, program=program,
-                bracket_map=bracket_map)
-
-        code = program[pc]
-
-        if code == ">":
-            tape.advance()
-
-        elif code == "<":
-            tape.devance()
-
-        elif code == "+":
-            tape.inc()
-
-        elif code == "-":
-            tape.dec()
-        
-        elif code == ".":
-            # print
-            os.write(1, chr(tape.get()))
-        
-        elif code == ",":
-            # read from stdin
-            tape.set(ord(os.read(0, 1)[0]))
-
-        elif code == "[" and tape.get() == 0:
-            # Skip forward to the matching ]
-            pc = get_matching_bracket(bracket_map, pc)
-            
-        elif code == "]" and tape.get() != 0:
-            # Skip back to the matching [
-            pc = get_matching_bracket(bracket_map, pc)
-
-        pc += 1
-
-class Tape(object):
-    def __init__(self):
-        self.thetape = [0]
-        self.position = 0
-
-    def get(self):
-        return self.thetape[self.position]
-    def set(self, val):
-        self.thetape[self.position] = val
-    def inc(self):
-        self.thetape[self.position] += 1
[…]

### 2015-05-07 laurens/ast/ast.py
Fixup jit-can-enter

@@ -7,10 +7,12 @@ class AST(object):
 
 class ValAST(AST):
   _attrs_ = ['value']
+  _immutable_fields_ = ['value']
   def __init__(self):
     raise Exception('Cannot instantiate')
 
 class Program(object):
+  _immutable_fields_ = ['binds']
   def __init__(self, binds):
     self.binds = binds
 

### 2015-05-07 laurens/op.py
New annotations

@@ -79,6 +79,7 @@ class Eval(Op):
   def __str__(self):
     return "\n\tExpr - " + str(self.expr) + "\n\tEnv - " + str(self.env)
 
+  @jit.unroll_safe
   def step(self,config):
     code       = config.code
     arg_stack  = config.arg_stack
@@ -238,6 +239,7 @@ class Enter(Op):
   def __str__(self):
     return "Enter - " + str(self.target)
 
+  @jit.unroll_safe
   def step(self, config):
     closure   = config.heap.lookup(config.code.target)
     lf        = closure.lam
@@ -312,6 +314,7 @@ class ReturnCon(Op):
   def __str__(self):
     return "Return Constructor - " + str(self.constructor) + str(self.rands)
 
+  @jit.unroll_safe
   def step(self, config):
     if config.ret_stack.empty():
         if config.upd_stack.empty():
@@ -374,6 +377,7 @@ class ReturnInt(Op):
   def __str__(self):
     return "Return Int - " + str(self.value)
 
+  @jit.unroll_safe
   def step(self,config):
     debug("Int return")
 

### 2015-05-07 laurens/op.py
Massive speed gain

@@ -15,7 +15,7 @@ from data.updframe import UpdateFrame
 from debug         import logMsg, debug
 from parse         import parse
 
-from rpython.rlib.jit import purefunction
+from rpython.rlib.jit import purefunction, unroll_safe 
 
 EvalOp      = 0
 EnterOp     = 1
@@ -79,7 +79,7 @@ class Eval(Op):
   def __str__(self):
     return "\n\tExpr - " + str(self.expr) + "\n\tEnv - " + str(self.env)
 
-  @jit.unroll_safe
+  @unroll_safe
   def step(self,config):
     code       = config.code
     arg_stack  = config.arg_stack
@@ -239,7 +239,7 @@ class Enter(Op):
   def __str__(self):
     return "Enter - " + str(self.target)
 
-  @jit.unroll_safe
+  @unroll_safe
   def step(self, config):
     closure   = config.heap.lookup(config.code.target)
     lf        = closure.lam
@@ -314,7 +314,7 @@ class ReturnCon(Op):
   def __str__(self):
     return "Return Constructor - " + str(self.constructor) + str(self.rands)
 
-  @jit.unroll_safe
+  @unroll_safe
   def step(self, config):
     if config.ret_stack.empty():
         if config.upd_stack.empty():
@@ -377,7 +377,7 @@ class ReturnInt(Op):
   def __str__(self):
     return "Return Int - " + str(self.value)
 
-  @jit.unroll_safe
+  @unroll_safe
   def step(self,config):
     debug("Int return")
 

### 2015-05-07 test/Fib.hs

@@ -1,7 +1,0 @@
-module Fib where
-
-fib 0 = 0
-fib 1 = 1
-fib n = fib (n - 1) + fib (n - 2)
-
-main = fib 30

### 2015-08-31 laurens/ast/ast.py
Refactored the eval out of its case

@@ -1,5 +1,36 @@
 from collections import deque
 
+@purefunction
+def vals(env, global_env, k):
+  logMsg("Vals with ", str(k))
+  if k == []:
+    return []
+  return [val(env, global_env, v) for v in k] # map(lambda v : val(env, global_env, v), k)
+
+@purefunction
+def val(env, global_env, k):
+  logMsg("Val with ", str(k))
+  res = k
+  if type(k) is ast.ast.Lit:
+    res = ast.ast.Value(k.value, True)
+  elif type(k) is ast.ast.Var:
+    var = k.variable
+    if var in env:
+      res = env[var]
+    elif var in global_env:
+      res = global_env[var]
+  elif type(k) is ast.ast.Atom:
+    litHuh = k.isLit
+    if litHuh:
+      res = ast.ast.Value(k.value, True)
+    elif k.value in env:
+      res = env[k.value]
+    else:
+      res = global_env[k.value]
+
+  assert isinstance(res,ast.ast.ValAST)
+  return res 
+
 class AST(object):
   _attrs_ = []
   def __init__(self):
@@ -48,7 +79,6 @@ class Lambda(AST):
     res += "\n\t - expr: "
     res += str(self.expr)
     return res
-  
 
 class Let(AST):
   def __init__(self, bindings, expr):
@@ -66,11 +96,54 @@ class Case(AST):
     self.case_expr = expr
     self.alts      = alts
 
+  def eval(self, config):
+    ret_stack  = config.ret_stack
+    cenv       = config.code.env
+    
+    debug("=> Case")
+    case      = self
+    local_env = cenv.copy()
+    config.ret_stack.push(ast.cont.CaseCont(case.alts, local_env))
+
+    config.code = op.Eval(case.case_expr, local_env)
+    return config
+
 class App(AST):
   def __init__(self, var, atoms):
     self.rator = var
     self.rands = atoms
- 
+
+  def eval(self, config):
+    cenv       = config.code.env
+    arg_stack  = config.arg_stack
+    global_env = config.global_env
+    
+    debug("=> App")
+    lookup   = val(cenv, global_env, self.rator)
+    lookupTy = type(lookup)
+    logMsg("Lookup", str(lookup))
+    if lookupTy is ast.ast.Value:
+      if lookup.isAddr:
+        logMsg("Rands", str(self.rands))
+        lookup_rands = vals(cenv, global_env, self.rands)
+
[…]

### 2015-08-31 laurens/op.py

@@ -23,37 +23,6 @@ ReturnConOp = 2
 ReturnIntOp = 3
 
 @purefunction
-def vals(env, global_env, k):
-  logMsg("Vals with ", str(k))
-  if k == []:
-    return []
-  return [val(env, global_env, v) for v in k] # map(lambda v : val(env, global_env, v), k)
-
-@purefunction
-def val(env, global_env, k):
-  logMsg("Val with ", str(k))
-  res = k
-  if type(k) is ast.ast.Lit:
-    res = ast.ast.Value(k.value, True)
-  elif type(k) is ast.ast.Var:
-    var = k.variable
-    if var in env:
-      res = env[var]
-    elif var in global_env:
-      res = global_env[var]
-  elif type(k) is ast.ast.Atom:
-    litHuh = k.isLit
-    if litHuh:
-      res = ast.ast.Value(k.value, True)
-    elif k.value in env:
-      res = env[k.value]
-    else:
-      res = global_env[k.value]
-
-  assert isinstance(res,ast.ast.ValAST)
-  return res 
-
-@purefunction
 def lookup_op(op):
   if (op == EvalOp):
     return "Eval"
@@ -81,155 +50,7 @@ class Eval(Op):
 
   @unroll_safe
   def step(self,config):
-    code       = config.code
-    arg_stack  = config.arg_stack
-    ret_stack  = config.ret_stack
-    upd_stack  = config.upd_stack
-    global_env = config.global_env
-    heap       = config.heap
-
-    cexp       = code.expr
-    cenv       = code.env
-    expr_type  = type(cexp)
-
-    if expr_type is ast.ast.App:
-      debug("=> App")
-      lookup   = val(cenv, global_env, cexp.rator)
-      lookupTy = type(lookup)
-      logMsg("Lookup", str(lookup))
-      if lookupTy is ast.ast.Value:
-        if lookup.isAddr:
-          logMsg("Rands", str(cexp.rands))
-          lookup_rands = vals(cenv, global_env, cexp.rands)
-
-          logMsg("Looked up", str(lookup_rands))
-          arg_stack.extend(lookup_rands)
-          
-          config.code = op.Enter(lookup.value)
-          debug(str(arg_stack.peek()))
-        
-        elif lookup.isInt and cexp.rands == []:
-          config.code = op.ReturnInt(lookup.value)
-        
-        else:  
-          raise Exception('integer literal in application position', lookup.value)
-      
-      else:
[…]

### 2015-08-31 laurens/op.py
More idiomatic python

@@ -50,7 +50,7 @@ class Eval(Op):
 
   @unroll_safe
   def step(self,config):
-    return config.code.eval(config)
+    return expr.eval(config)
 
 class Enter(Op):
   def __init__(self,addr):
